Highly Available Transactions: Virtues and Limitations

### 2013-03-31 writeup/futurework.tex
References purge

\textbf{Where do semantics-based concurrency control and weakened
  failure models lie in the HAT landscape?} The vast amount of
literature on semantics-based concurrency control provides hints as to
when coordination can be avoided while providing ``strongly
consistent'' outcomes for limited failure scenarios. We have performed
a narrow case study in this work, but a general taxonomization of
techniques such as escrow transactions~\cite{escrow},
Sagas~\cite{sagas}, and other long-running transaction
support~\cite{gray-virtues} could help answer the challenge of
programming in a weakly consistent environment. HAT guarantees can
also be strengthened by considering weaker failure models such as
bounded asynchrony or models in which the underlying hardware provides
explicit service-level agreements. Additionally, while we have focused
on systems that provide a single availability mode, there are several
interesting hybrids to consider. For example, a system may provide
``best-effort'' stickiness (or serializability) in the absence of
partitions, but, in the presence of partitions, ``lose'' guarantees by
having clients reconnect to an available replica. This strategy
provides stronger guarantees in the absence of partitions but, to
provide availability, sacrifices them in the event of partitions.

>>>>

\textbf{Where do semantics-based concurrency control and weakened
  failure models lie in the HAT landscape?} The vast amount of
literature on semantics-based concurrency control provides hints as to
when coordination can be avoided while providing ``strongly
consistent'' outcomes for limited failure scenarios. We have performed
a narrow case study in this work, but a general taxonomization of
techniques such as escrow transactions~\cite{escrow},
Sagas~\cite{sagas}, and other long-running transaction support could
help answer the challenge of programming in a weakly consistent
environment. HAT guarantees can also be strengthened by considering
weaker failure models such as bounded asynchrony or models in which
the underlying hardware provides explicit service-level
agreements. Additionally, while we have focused on systems that
provide a single availability mode, there are several interesting
hybrids to consider. For example, a system may provide ``best-effort''
stickiness (or serializability) in the absence of partitions, but, in
the presence of partitions, ``lose'' guarantees by having clients
reconnect to an available replica. This strategy provides stronger
guarantees in the absence of partitions but, to provide availability,
sacrifices them in the event of partitions.

### 2013-03-31 writeup/hat-analysis.tex
References purge

HAT systems, or systems providing transactions with transactional
availability with high availability ($1$-available) or sticky high
availability ($1$-sticky-available), offer substantial latency and
availability benefits, yet they come with a cost to achievable
semantics. In this section, we delineate which of several ACID,
distributed consistency, and session consistency levels can be
achieved with high availability (transactional atomicity, variants of
Repeatable Read isolation, and many session guarantees) and sticky
high availability (read your writes, PRAM and causal consistency) and
which cannot (preventing Lost Update and Write Skew, recency).  We
present a full summary of these results in
Section~\ref{sec:hat-summary}.

As Brewer states, ``systems and database communities are separate but
overlapping (with distinct vocabulary)''~\cite{brewer-slides}. With
this challenge in mind, we draw on existing properties and definitions
from the database and distributed systems literature, providing a
brief, informal explanation and example for each guarantee. In
particular, for our ACID guarantees, we draw largely on Atul Adya's
dissertation~\cite{adya}, Berenson et al.'s 1995 critique of the ANSI
SQL Specification~\cite{ansicritique}, and the ANSI SQL
specification~\cite{ansi-sql}. We provide a set of formal definitions
and semantics in our extended Technical Report~\cite{hat-tr}.

@@

To begin, we present achievable semantics, offering proof-of-concept
algorithms to demonstrate feasibility. These guarantees are fairly
straightforward and have been introduced---albeit with greater brevity
and discussion omitted---in a preliminary workshop
paper~\cite{hat-hotos}. In our examples, we exclusively consider read
and write operations. We denote a write of value $v$ to data item $d$
as $w_d(v)$ and a read from data item $d$ returning $v$ as $r_d(v)$
and assume that all data items have the null value, $\bot$, at
database initialization. Unless otherwise specified, all example
transactions commit.

@@

To begin, Adya's \textit{Read Uncommitted} isolation requires a total
order among transactions and requires that each transaction's writes
are ordered consistently with this order (prohibiting ``Dirty
Writes,'' or $G0$)~\cite{adya}. If two transactions write to the same
set of data items, then the final database state cannot contain the
``earlier'' transaction's writes to the data items. For example, in
the below example, $T_3$ should eventually only read $a=b=1$ or
$a=b=2$ but not $a=2, b=1$ or $a=1, b=2$:
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)\\[-2em]
\end{align*}
We will strengthen this property in later models, which will prevent,
for instance, reading $a=2, b=1$ at any time (not just
``eventually''). The difference between this total order on
transactions and the total order required by serializability is that
the Read Uncommitted total order is completely arbitrary (i.e., an
order must exist), whereas a serializable total order must be
equivalent to a serial execution. Read Uncommitted is easily achieved
via applying the same logical timestamp to each update in a
transaction and applying a ``last writer wins'' conflict
reconciliation policy at each replica.

\textit{Read Committed} isolation requires that transactions do not
read uncommitted versions of data items (prohibiting both ``Dirty
Writes''---as above---and ``Dirty Reads'' phenomena; captured by
Adya's $G1a-c$, ANSI's $P1$, and ``broad'' $P1$ (2.2) from Berenson et
al.). For instance, in the example below, $T_3$ should never see
$a=1$, and, if $T_2$ aborts, $T_3$ should never read $a=3$:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\\[-2em]
\end{align*}
It is fairly easy to prvent ``Dirty Reads'': if each transaction never
writes uncommitted data to the database, then transactions will never
read each others' dirty data. As a simple solution, clients can buffer
their writes until they commit, or, alternatively, can send them to
servers, who will not serve new writes until clients notify them that
the writes have been committed.

\textit{Repeatable Read} isolation is a contentious property. As
Berenson et al. discuss~\cite{ansicritique}, Gray's original
Repeatable Read lock-based implementation provides substantially
richer guarantees than those that are required by the ANSI SQL
specification~\cite{gray-isolation}. Gray~\cite{gray-isolation},
Berenson et al.~\cite{ansicritique}, and Adya~\cite{adya} all
interpret Repeatable Read as providing serializability for all
operations except predicate-based reads. However, in its description
of Repeatable Read, the ANSI SQL specification provides a useful
property which, although is not true to Gray's spirit of Repeatable
Read, is also commonly found in distributed consistency models: the
ANSI Repeatable Read guarantee requires that, along with observing
Read Committed isolation, each transaction will only read one version
of each data item that it did not itself produce (preventing ``Fuzzy
Read,'' or P2). In the example below, $T_3$ must read $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This isolation property is a literal interpretation of the phrase
``Repeatable Read'': unless a transaction modifies a given data item,
the observable value of the data item should not change during the
transaction. By itself, this property is rather weak---we can always
read $\bot$ for each data item. However, when coupled with additional
properties, like transactional atomicity, it is stronger: the repeated
reads must obey additional ordering constraints
(Section~\ref{sec:discussion}). In fact, distributed systems often
describe a ``consistent snapshot'' across a set of related events as a
\textit{cut} across a set of participants or, in our case, data
items. Accordingly, to capture the notion that the transaction should
read from a (non-changing) consistent cut over data items (and to
disambiguate from the aforementioned stronger Repeatable Read
properties), we call ANSI SQL ``Repeatable Read'' (i.e., preventing
P2) \textbf{Cut Isolation}. It is possible to satisfy cut isolation
with high availability by caching appropriate versions for each of a
transaction's reads or, alternatively, using multi-versioning on each
replica and ensuring that each of a transaction's successive reads
return the same version of each data item.

We can further consider two variants of Cut Isolation. The first is
relatively straightforward and provides a cut across individual data
items. We will call this \textit{Item Cut Isolation}. The second
provides a cut over individual data items and over logical ranges of
data items, or predicate-based reads (e.g., \texttt{SELECT
  WHERE}). This prevents the \textit{Phantom Problem}, whereby two
successive predicate-based reads return different
data~\cite{gray-isolation}. We call predicate-based cut isolation,
which prevents Phantoms, \textit{Predicate Cut
  Isolation}.\footnote{Oracle provides an isolation level called
  ``Statement Level Read Consistency''; this is analogous to Predicate
  Cut Isolation at the level of a single operation within a
  multi-operation transaction. We do not consider this isolation model
  here as it is subsumed by standard Snapshot Isolation and we believe
  our discussion is easily extended to incorporate it.}

@@

Transactional atomicity (TA) is core to ACID guarantees. Although, at
least by the ACID acronym, it is not an ``isolation'' property,
transactional atomicity restricts transactions' ability to view the
effects of partially completed transactions. Under transactional
atomicity, within each transaction, either all effects of another
transaction are observed, or none are (equivalently, once some of the
effects of a transaction are observed, all effects are observed). This
is a strictly stronger guarantee than Read Committed isolation: in
Read Committed, we can read a subset of another transaction's
committed writes whereas, with TA, we must read all or none of
them. Together with item cut isolation, TA prevents Read Skew
anomalies (Berenson et al.'s A5A~\cite{ansicritique}). As an example
of TA, $T_2$ must observe $b=c=1$ (or later versions for each key):
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$, $a=1$, or a later version of
$a$. Notably, TA requires Read Committed isolation: observing all
effects of a transaction implicitly requires observing the final
(committed) effects of a transaction as well.

Perhaps perplexingly, discussions of TA are absent from existing
discussions of weak isolation. This is perhaps again due to the
single-node context in which prior work was developed: on a single
server (or a fully replicated database), TA is achievable via
lightweight locking and/or local concurrency control over data
items. In contrast, in a distributed environment, TA over arbitrary
groups of non-colocated items is considerably more difficult to
achieve with high availability: servers need to know when all of a
transaction's updates are present on their respective
replicas. However, replicas do not need to agree on when to reveal a
new value to clients, which would require consensus; accordingly, TA
only requires reliable broadcast (with some additional metadata for
clients) and is achievable in a HAT system. We omit a full discussion
of the algorithm and instead refer the reader to the extended version
of this paper~\cite{hat-tr}.

@@

Our models have not yet considered interactions \textit{across}
transactions: we have not guaranteed any ordering or continuity
between transactions (other than that there exists some arbitrary
ordering). This is a natural concern in a distributed context, and, in
the distributed systems literature, many useful \textit{safety}
guarantees span multiple application actions. In particular,
\textit{session guarantees} are used to describe guarantees across
transactions within a given \textit{session}, ``an abstraction for the
sequence of...operations performed during the execution of an
application''~\cite{sessionguarantees}. Informally a session describes
a context that should persist between transactions: for example, on a
social networking site, all of a user's transactions submitted between
``log in'' and ``log out'' operations might form a session.

@@

\vspace{.5em}\noindent\textit{{Monotonic reads}} requires that,
within a session, subsequent reads to a given object ``never return any
previous values''; reads from each item progress forward in a total
order. The ordering of reads should respect any total ordering on
transactions.

\vspace{.5em}\noindent\textit{{Monotonic writes}} requires that
each session's writes be serialized. Any order on transactions should
also respect the order in which each session submitted transactions.

\vspace{.5em}\noindent\textit{{Writes Follow Reads}} requires that, if
a session observes an effect of transaction $T_2$ and subsequently
commits transaction $T_2$, then another session can only observe
effects of $T_2$ if it can also observe $T_1$'s (or later values for
$T_1$'s effects). Any order on transactions should respect the
reads-from order.\vspace{.5em}

We can achieve the above guarantees by forcing servers to wait to
reveal new writes until each write's respective dependencies are
fulfilled on all replicas. This mechanism effectively ensures that all
clients read from a globally agreed upon lower bound on the versions
written to the system. This \textit{is} highly available as a client
will never block due to inability to find a server with a sufficiently
up-to-date version of a data item. However, it does not imply that
transactions will read their own writes or, in the presence of
partitions, make forward progress through the version history. The
problem is that, if a server becomes partitioned, under the highly
available model, we must handle the possibility that an unfortunate
client will be forced to issue her next requests against the
partitioned server!

@@

\vspace{.5em}\noindent\textit{{Causal
    consistency}}~\cite{causalmemory} is the combination of all of the
session guarantees~\cite{sessiontocausal} (alternatively, PRAM with
writes-follow-reads) and is also referred to as PL-2L
isolation~\cite{adya}).\vspace{.5em}

Read your writes is not achievable in a highly available
($1$-available) system. Consider a client that executes the following
two transactions in succession:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: r_x(a)
\end{align*}
If the client executes $T_1$ against a server that is partitioned from
the rest of the other servers, the server must allow $T_1$ to
commit. If the client executes $T_2$ against the same (partitioned)
server, then it will be able to read its writes. However, if the
network topology shifts and the client can only contact a different
server which is partitioned from the server that executed $T_1$, then
the client will be unable to read its own writes and the system will
have to either stall indefinitely to allow the client to read her
writes (violating transactional availability) or will have to
sacrifice read your writes guarantees. However, if the client remains
sticky with the server that executed $T_1$, then we can disallow this
scenario. Accordingly, read your writes, and, by proxy, causal
consistency and PRAM require stickiness. Read your writes is ``free''
in a sticky system, while the remaining causality and PRAM guarantees
can be accomplished with the algorithms for achieving the remaining
session guarantees.

>>>>

HAT systems provide transactions with transactional availability and
either high availability ($1$-available) or sticky high availability
($1$-sticky-available). They offer substantial latency and
availability benefits compared to traditional distributed databases,
yet they cannot achieve all the traditional semantics. In this
section, we delineate ACID, distributed consistency, and session
consistency levels which can be achieved with high availability
(transactional atomicity, variants of Repeatable Read isolation, and
many session guarantees), those with sticky high availability (read
your writes, PRAM and causal consistency) and the properties which
cannot be provided in a HAT system (those preventing Lost Update and
Write Skew, or with recency).  We present a full summary of these
results in Section~\ref{sec:hat-summary}.

As Brewer states, ``systems and database communities are separate but
overlapping (with distinct vocabulary)''~\cite{brewer-slides}. With
this challenge in mind, when possible, we build on existing properties
and definitions from the database and distributed systems literature,
providing a brief, informal explanation and example for each
guarantee. The database isolation guarantees require particular care,
since different DBMSs often use the same terminology for different
mechanisms and may provide additional guarantees in addition to our
implementation-agnostic definitions.  We draw largely on Atul Adya's
dissertation~\cite{adya} and somewhat on its predecessor work: the
ANSI SQL specification~\cite{ansi-sql} and Berenson et al.'s
subsequent 1995 critique~\cite{ansicritique}. We provide a set of
formal definitions, semantics, proofs and observations in our extended
Technical Report and opt for a more informal presentation in this
paper~\cite{hat-tr}.

@@

To begin, we present well-known semantics that can be achieved in HAT
systems. We offer proof-of-concept highly available algorithms but our
goal is \textit{not} to provide optimal or even efficient
implementations. These guarantees are fairly straightforward and have
been introduced---albeit with greater brevity and discussion
omitted---in a preliminary workshop paper~\cite{hat-hotos}. In our
examples, we exclusively consider read and write operations. We denote
a write of value $v$ to data item $d$ as $w_d(v)$ and a read from data
item $d$ returning $v$ as $r_d(v)$ and assume that all data items have
the null value, $\bot$, at database initialization. Unless otherwise
specified, all example transactions commit.

@@

To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
PL-1,requiring that each transaction's writes are ordered consistently
according to a single total order on transactions. This prohibits
Adya's phenomenon $G0$, also called ``Dirty Writes''~\cite{adya}. If
two transactions write to the same set of data items, then the final
database state cannot contain the ``earlier'' transaction's writes to
the data items. For example, in the below example, $T_3$ should
\emph{eventually} only read $a=b=1$ or $a=b=2$ but not $a=2, b=1$ or
$a=1, b=2$:
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)\\[-2em]
\end{align*}
All later properties will strengthen PL-1.

Unlike the total order required by serializability, the order in PL-1
does not constrain the values seen in by a transaction's read
operations except once the database has reached a ``final
state.'' Read Uncommitted is easily achieved via applying the same
logical timestamp to each update in a transaction and applying a
``last writer wins'' conflict reconciliation policy at each replica.

\textbf{Read Committed} isolation is particularly important in
practice as it is the default of many DBMSs. Implementations differ,
with some based on long-duration exclusive locks and short-duration
read locks~\cite{gray-isolation} and others based on multiple
versions. The implementations typically have recency and monotonicity
properties beyond the simple meaning of the name, which is what is
expressed in the implementation-agnostic definition: under Read
Committed, transactions should do not access uncommitted or
intermediate versions of data items. This prohibits both ``Dirty
Writes'', as above, and also ``Dirty Reads'' phenomena.  This
isolation is Adya's PL-2, and formalised by prohibiting Adya's $G1a-c$
(or ANSI's $P1$, or ``broad'' $P1$ (2.2) from Berenson et al.). For
instance, in the example below, $T_3$ should never see $a=1$, and, if
$T_2$ aborts, $T_3$ should never read $a=3$:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\\[-2em]
\end{align*}
It is fairly easy for a HAT system to prevent ``Dirty Reads'': if each
transaction never writes uncommitted data to the database, then
transactions will never read each others' dirty data. As a simple
solution, clients can buffer their writes until they commit, or,
alternatively, can send them to servers, who will not deliver their
value to other readers until notified that the writes have been
committed. This implementation does not provide recency or
monotonicity guarantees but satisfies the implementation-agnostic
definition.

``Repeatable Read'' isolation is a confusing property. Some IBM
products use the term for fully serializable
isolation~\cite{hat-hotos}. Gray~\cite{gray-isolation}, Berenson et
al.~\cite{ansicritique}, and Adya's PL-2.99~\cite{adya} all interpret
Repeatable Read as providing serializability for all operations except
predicate-based reads (and so Repeatable Read is identical to
serializability in a key-value store without predicate-based
access). In this model, the \textit{Phantom Problem}, whereby two
successive predicate-based reads return different
data~\cite{gray-isolation}, is the only kind of non-serializable
behavior allowed. When referring to ``Repeatable Read,'' we will
follow these definitions, which, as we will soon show, are for a
property that be offered in a highly available way.

However, As Berenson et al.~\cite{ansicritique} discuss, the ANSI SQL
specification allows many additional behaviors under the term
Repeatable Read isolation.  \textbf{ANSI Repeatable Read} requires
that, along with respecting Read Committed isolation, each transaction
will only read one version of each data item that it did not itself
produce (preventing ``Fuzzy Read,'' or P2). In the example below,
$T_3$ must read $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This is a literal interpretation of the phrase ``Repeatable Read'':
unless a transaction modifies a given data item, the observable value
of the data item should not change during the transaction. By itself,
this property is extremely weak---we can always read $\bot$ for each
data item---and easily achieved in a HAT system by caching the values
read.

Between our Repeatable Read (PL-2.99) and the ANSI definition, is a
concept we will call \textbf{Item Cut Isolation}: that all the
transaction's reads should see values from a non-changing consistent
cut or snapshot over the data items. It is possible to satisfy Item
Cut Isolation with high availability by using multi-versioning on each
replica and ensuring that each of a transaction's reads return an
appropriately chosen version (e.g., identified by vector clock).

A stronger achievable property asks that a consistent cut spans both a
transaction's predicate reads (e.g., \texttt{SELECT WHERE}) as well as
its item reads.  We call this \textit{Predicate Cut
  Isolation}\footnote{Oracle provides an isolation level called
  ``Statement Level Read Consistency''~\cite{adya}; this is analogous
  to Predicate Cut Isolation at the level of a single operation within
  a multi-operation transaction. We do not consider this isolation
  model here as it is subsumed by standard Snapshot Isolation and we
  believe our discussion is easily extended to incorporate it.} and it
prohibits Phantoms. Predict Cut Isolation is also achievable in HAT
system via similar multi-versioning as above.

@@

Transactional atomicity (TA) is core to ACID guarantees. Although, at
least by the ACID acronym, it is not an ``isolation'' property,
transactional atomicity restricts the ability to view the effects of
partially completed transactions. Under transactional atomicity, once
some of the effects of a transaction $T_1$ are observed by another
transaction $T_2$, thereafter all effects of $T_1$ are observed by
$T_2$. Together with item cut isolation, TA prevents Read Skew
anomalies (Berenson et al.'s A5A~\cite{ansicritique}). As an example
of TA, because $T_2$ has read $T_1$'s write to $y$, $T_2$ must observe
$b=c=1$ (or later versions for each key):
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$, $a=1$, or a later version of
$a$. Notably, TA requires Read Committed isolation: observing all
effects of a transaction implicitly requires observing the final
(committed) effects of a transaction as well.

Perplexingly, discussions of TA are absent from existing discussions
of weak isolation. This is perhaps again due to the single-node
context in which prior work was developed: on a single server (or a
fully replicated database), TA is achievable via lightweight locking
and/or local concurrency control over data items~\cite{gstore}. In
contrast, in a distributed environment, TA over arbitrary groups of
non-colocated items is considerably more difficult to achieve with
high availability: servers need to know when all of a transaction's
updates are present on their respective replicas. However, replicas do
not need to agree on when to reveal a new value to clients, which
would require consensus; accordingly, TA only requires reliable
broadcast (with some additional metadata for clients) and is
achievable in a HAT system. We omit a full discussion of the algorithm
and instead refer the reader to the extended version of this
paper~\cite{hat-tr}.  space{.5em}

@@

Our models have not yet considered interactions \textit{across}
transactions: we have not guaranteed any ordering or continuity
between transactions, other than that there exists some arbitrary
ordering (via Read Uncommitted). In the distributed systems
literature, many useful \textit{safety} guarantees span multiple
application actions. In particular, \textit{session guarantees} are
used to describe guarantees across transactions within a given
\textit{session}, ``an abstraction for the sequence of...operations
performed during the execution of an
application''~\cite{sessionguarantees}. Informally a session describes
a context that should persist between transactions: for example, on a
social networking site, all of a user's transactions submitted between
``log in'' and ``log out'' operations might form a session.

@@

\vspace{.5em}\noindent\textit{{Monotonic reads}} requires that, within
a session, subsequent reads to a given object ``never return any
previous values''; reads from each item progress forward in a total
order. The ordering of reads should respect any externally observable
total ordering on transactions.

\vspace{.5em}\noindent\textit{{Monotonic writes}} requires that each
session's writes be applied to any replica in the order they were
submitted by the client. Any order on transactions should also be
consistent with any precedence that a global observer would see.

\vspace{.5em}\noindent\textit{{Writes Follow Reads}} requires that, if
a session observes an effect of transaction $T_1$ and subsequently
commits transaction $T_2$, then another session can only observe
effects of $T_2$ if it can also observe $T_1$'s effects (or later
values that supercede $T_1$'s).  Any order on transactions should
respect the reads-from order.\vspace{.5em}

We can achieve the above guarantees by forcing servers to wait to
reveal new writes until each write's respective dependencies are
fulfilled on all replicas. This mechanism effectively ensures that all
clients read from a globally agreed upon lower bound on the versions
written to the system. This \textit{is} highly available as a client
will never block due to inability to find a server with a sufficiently
up-to-date version of a data item. However, it does not imply that
transactions will read their own writes or, in the presence of
partitions, make forward progress through the version history. The
problem is that, if a server becomes partitioned, under the highly
available model, we must handle the possibility that an unfortunate
client will be forced to issue her next requests against the
partitioned server.

@@

\vspace{.5em}\noindent\textit{{Causal
    consistency}}~\cite{causalmemory} is the combination of all of the
session guarantees~\cite{sessiontocausal} (alternatively, PRAM with
writes-follow-reads) and is also referred to by Adya as PL-2L
isolation~\cite{adya}).\vspace{.5em}

Read your writes is not achievable in a highly available
($1$-available) system. Consider a client that executes the following
two transactions in succession:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: r_x(a)
\end{align*}
If the client executes $T_1$ against a server that is partitioned from
the rest of the other servers, the server must allow $T_1$ to
commit. If the client executes $T_2$ against the same (partitioned)
server, then it will be able to read its writes. However, if the
network topology shifts and the client can only contact a different
server that is partitioned from the server that executed $T_1$, then
the client will be unable to read its own writes and the system will
have to either stall indefinitely to allow the client to read her
writes (violating transactional availability) or will have to
sacrifice read your writes guarantees. However, if the client remains
sticky with the server that executed $T_1$, then we can disallow this
scenario. Accordingly, read your writes, and, by proxy, causal
consistency and PRAM require stickiness. Read your writes is ``free''
in a sticky system, while the remaining causality and PRAM guarantees
can be accomplished with the algorithms for achieving the remaining
session guarantees.

### 2013-03-31 writeup/intro.tex
PDB pass plus Alan comments
References purge

The prior decade saw a large shift in the design of large-scale
database backends, from the use of transactional
RDBMSs~\cite{bernstein-concurrency, bernstein-book, gray-isolation,
  gray-virtues} to the widespread adoption of loosely-consistent
distributed key-value stores~\cite{bigtable, pnuts,
  dynamo}. Instrumental in this trend was the 2000 introduction of
Brewer's CAP Theorem, which stated that a highly available system
cannot provide strong consistency guarantees in the presence of
network partitions~\cite{brewer-slides}. However, as formally
proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability~\cite{herlihy-art}, the ability to read the most
recent write to a data item that is replicated across servers. Yet
this result---regarding a fairly narrow form of distributed
consistency---is often conflated with ACID database
properties~\cite{brewer-slides, hn, foundation-article}, leading to
substantial confusion about the relationships between replica
consistency, transactional isolation, and high availability. However,
transactions \textit{are} useful for programmers as evidenced by the
recent resurgence of transactional data stores such as Google's
Spanner~\cite{spanner}. Unfortunately, these systems are not highly
available in the presence of partitions~\cite{walter, mdcc,
  foundation-article, krikellas-bargain, eiger}. (TODO:ADD more)

@@

Unfortunately, the relationship of the CAP Theorem and ACID semantics
has not been well explored. We have a strong understanding of weak
isolation in the single-server context from which it
originated~\cite{adya, ansicritique, gray-isolation} and many papers
offer techniques for providing distributed
serializability~\cite{bernstein-concurrency, spanner, granola,
  daudjee-session, krikellas-bargain, calvin} or snapshot
isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
  walter}. The distributed computing, programming language, and
parallel hardware literature also contains many consistency models for
single operations on replicated objects~\cite{pnuts, herlihy-art,
  cops, eiger, cac, sessionguarantees}. However, the literature has
largely ignored semantic guarantees for multiple operations operating
on multiple data items in a highly available distributed environment.

Our main contributions are as follows. We fill gap between, on one
hand, the plethora of previously proposed isolation and consistency
models, and, on the other hand, the goal of high availability.  We
classify which, among the wide array models, are highly available,
denoting them as {\em Highly Available Transactions (HATs)}. In doing
so, we demonstrate that many \textit{implementations} of HAT semantics
are not highly available but that their unavailability is not
\textit{inherently} required. Our investigation shows, somewhat
surprisingly, that besides serializability, only Snapshot Isolation
(SI) and Repeatable Reads (RR) are unavailable, with most other
isolation levels being available. We also show that many of the weaker
data consistency models from distributed systems are HAT-compliant.

More precisely, the reason why fairly strong models like Snapshot
Isolation (SI) and Repeatable Reads (RR) are not highly available is
that they require detecting conflicts between concurrent updates (as
needed for preventing Lost Updates or Write Skew phenomena). However,
Read Committed isolation, Transactional Atomicity, and many of the
weaker data consistency models from database and distributed systems
can be implemented in a highly available manner. For several of these
guarantees, such as Transactional Causal Predicate Cut Isolation, we
consider a modified form of high availability in which clients ``stick
to'' (i.e., have affinity with) at least one server (a property which
is often implicit in the distributed systems
literature~\cite{herlihy-art, cops, eiger, cac} but which requires
\textit{explicit} consideration in a client-server replicated database
context). This sticky availability is easily implementable and
widely employed~\cite{cops, eiger, vogels-defs} but is strictly less
available than traditional definitions of high availability.

>>>>

The prior decade saw a large shift in the design of large-scale
database backends, from the use of transactional
RDBMSs~\cite{bernstein-book, gray-isolation,
  gray-virtues} to the widespread adoption of loosely-consistent
distributed key-value stores~\cite{bigtable, pnuts,
  dynamo}. Instrumental in this trend was the 2000 introduction of
Brewer's CAP Theorem, which stated that a highly available system
cannot provide strong consistency guarantees in the presence of
network partitions~\cite{brewer-slides}. However, as formally
proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability~\cite{herlihy-art}, the ability to read the most
recent write to a data item that is replicated across servers. Yet
this result---regarding a fairly narrow form of distributed
consistency---is often conflated with ACID database
properties~\cite{brewer-slides, hn, foundation-article}, leading to
substantial confusion about the relationships between replica
consistency, transactional isolation, and high availability. However,
transactions \textit{are} useful for programmers as evidenced by the
recent resurgence of transactional data stores such as Google's
Spanner~\cite{spanner}. Unfortunately, these systems are not highly
available in the presence of partitions~\cite{walter, mdcc,
  foundation-article, krikellas-bargain, eiger}. (TODO:ADD more)

@@

Unfortunately, the relationship of the CAP Theorem and ACID semantics
has not been well explored. We have a strong understanding of weak
isolation in the single-server context from which it
originated~\cite{adya, ansicritique, gray-isolation} and many papers
offer techniques for providing distributed
serializability~\cite{bernstein-book, hstore, kemme-classification,
  spanner, daudjee-session, krikellas-bargain, calvin} or
snapshot isolation~\cite{daudjee-snapshot,generalizedsnapshot,
  kemme-snapshot, walter}. The distributed computing, programming
language, and parallel hardware literature also contains many
consistency models for single operations on replicated
objects~\cite{pnuts, herlihy-art,  eiger, cac, sessionguarantees,
  ceri-mechanism, chen-mechanism}. However, the literature has largely
ignored semantic guarantees for multiple operations operating on
multiple data items in a highly available distributed environment.

Our main contributions are as follows. We fill gap between, on one
hand, the plethora of previously proposed isolation and consistency
models, and, on the other hand, the goal of high availability.  We
classify which, among the wide array of models, are highly available,
denoting them as {\em Highly Available Transactions (HATs)}. In doing
so, we demonstrate that many \textit{implementations} of HAT semantics
are not highly available but that their unavailability is not
\textit{inherently} required. Our investigation shows, somewhat
surprisingly, that besides serializability, only Snapshot Isolation
(SI) and Repeatable Reads (RR) are unavailable, with most other
isolation levels being available. We also show that many of the weaker
data consistency models from distributed systems are HAT-compliant.

More precisely, the reason why fairly strong models like Snapshot
Isolation (SI) and Repeatable Reads (RR) are not highly available is
that they require detecting conflicts between concurrent updates (as
needed for preventing Lost Updates or Write Skew phenomena). However,
Read Committed isolation, Transactional Atomicity, and many of the
weaker data consistency models from database and distributed systems
can be implemented in a highly available manner. For several of these
guarantees, such as Transactional Causal Predicate Cut Isolation, we
consider a modified form of high availability in which clients ``stick
to'' (i.e., have affinity with) at least one server (a property which
is often implicit in the distributed systems
literature~\cite{herlihy-art,  eiger, cac} but which requires
\textit{explicit} consideration in a client-server replicated database
context). This sticky availability is easily implementable and
widely employed~\cite{eiger, vogels-defs} but is strictly less
available than traditional definitions of high availability.

### 2013-03-31 writeup/model.tex
References purge

Traditionally, a system provides high availability if every user that
can contact a server eventually receives a response from that server,
even in the presence of arbitrary, indefinitely long network
partitions between servers~\cite{gilbert-cap}. As in a standard
distributed database, designated servers might handle operations on
different data items;\footnote{There is a further distinction between
  a \textit{fully replicated} system, in which all servers are
  replicas for all data items, and a \textit{partially replicated}
  system, in which at least one server acts as a replica for only a
  subset of all data items. For generality, and, for applicability to
  many modern ``sharded'' or ``partitioned'' data storage
  systems~\cite{dynamo, pnuts, bigtable, spanner}, we consider partially
  replicated systems.} a server that can handle an operation for a
given data item is called a \textit{replica} for that item. While this
standard model is useful, there are several other cases we should
consider.

>>>>

Traditionally, a system provides high availability if every user that
can contact a server eventually receives a response from that server,
even in the presence of arbitrary, indefinitely long network
partitions between servers~\cite{gilbert-cap}. As in a standard
distributed database, designated servers might handle operations on
different data items;\footnote{There is a further distinction between
  a \textit{fully replicated} system, in which all servers are
  replicas for all data items, and a \textit{partially replicated}
  system, in which at least one server acts as a replica for only a
  subset of all data items. For generality, and, for applicability to
  many modern ``sharded'' or ``partitioned'' data storage
  systems~\cite{dynamo, pnuts, bigtable, spanner, hstore}, we consider partially
  replicated systems.} a server that can handle an operation for a
given data item is called a \textit{replica} for that item. While this
standard model is useful, there are several other cases we should
consider.
